---
title: "WI Stream Classes: exploratory data analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

## Objectives / To Do

+ classify/cluster streams by fish community & environmental data
* run LDA to predict and check classification accuracy
+ calculate fish metrics among classes = standards / opportunities


```{r setup, include=FALSE}
ragg_png = function(..., res = 150) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, 
                      dev = "ragg_png", res = 500, retina = 1, 
                      fig.align = 'center')
```

```{r prep}
# packages
library(tidyverse)  # data manipulation
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
library(patchwork)  # panel plots
library(vegan)      # community analsyses
library(pdftools)   # covert pdfs to pngs

options(ggrepel.max.overlaps = Inf)


# function to do kmeans analysis on a data given a # of clusters
f_kmeans <- function(dat, clusters, ...){
  kmeans(x = dat, centers = clusters, nstart = 25)
  }

# function to get total within-cluster sum of square 
wss <- function(kmeans_ob) {
  kmeans_ob$tot.withinss
  }
  
f_get_clusters <- function(cluster_ob, ...){
  cluster_ob$cluster 
}

# function to plot wss
f_wss_plot <- function(k, kmeans_ob, ...){
  # make df
  pdat <- tibble(
    clusters = 1:k, 
    wss = map_dbl(kmeans_ob, wss)
    ) 
  # plot
  p.wss <- pdat %>% 
    ggplot(aes(clusters, wss)) +
    geom_point() + 
    geom_line() + 
    scale_x_continuous(breaks = seq(0,10,1)) +
    theme_minimal() + 
    labs(
      x="Number of clusters (K)", 
      y="WSS", 
      title = "Elbow test for optimal clusters"
      )
}

# function to plot kmeans cluster objects
f_fviz_cluster_plots <- function(dat, kmeans_ob, ...){
  fviz_cluster(
    object = kmeans_ob, 
    data = dat, 
    geom = "point", 
    main = paste("Clusters = ", length(kmeans_ob$size)),
    ggtheme = theme_minimal()
    )
}

# function to plot kmeans cluster on map
f_plot_maps <- function(stream_clusters, clusters){
  d1 <- stream_clusters %>% filter(k == clusters)
  d2 <- whd_lines %>% 
    left_join(d1, by="reach_id") %>% 
    filter(!is.na(group))
  ggplot() +
  geom_sf(data=wdnr.gis::wi_poly, color = "black", fill="white") + 
  geom_sf(data=whd_lines %>% filter(wbic==1179900), color="black") +
  geom_sf(data=d2, aes(color = as.factor(group)), size=1) +
  theme_void() + 
  theme(
    legend.position="none", 
    plot.margin = unit(c(0, 0, 0, 0), "null")
    )
}

```


## Data

```{r read-data}
source(here::here("R", "13_prep-data.R"))  # ~1-2 min load time
```

There are two datasets: 

* **Fish community**: Catch and effort data for all electrofishing surveys (boom, backpack, barge) done on streams by the WDNR from 1994-2020. Excludes all lake and Mississippi River sites. Data was pulled from the WDNR FMIS on 18 April 2022. 
* **24k WI Hydrography**: A geodatabase of channel, riparian, and watershed attributes for all stream and lake features in 1:24,000 WI Hydrography database. Available at for [download here](https://www.arcgis.com/home/item.html?id=c4bc634ba115498487174bda137f8de8).


## Fish Community

Partition streams into groups based on reach-level fish species presence-absence data. We do this for two objectives: 

* the whole fish community
* primary sportfish species (fisheries)

```{r spp-groups}
# list of sportfish species
sport.spp <- c(
  "muskellunge", "northern_pike", "walleye", "sauger",
    "largemouth_bass", "smallmouth_bass", 
    "channel_catfish", "flathead_catfish",
    "brook_trout", "brown_trout", "rainbow_trout",
    "lake_sturgeon", 
    "black_crappie", "bluegill", "green_sunfish", "pumpkinseed", 
    "rock_bass", "yellow_perch"
  )

# list of pan fish species (subset of above)
pan.spp <- c(
    "black_crappie", "bluegill", "green_sunfish", "pumpkinseed", 
    "rock_bass", "yellow_perch"
  )

# list of trout species  (subset of above)
trout.spp <- c("brook_trout", "brown_trout", "rainbow_trout")

# list of catfish spp (subset of above
catfish.spp <- c("channel_catfish","flathead_catfish")
```

### Species prevelance

```{r spp-prevelance, fig.height=16, fig.width=8}
spp_prev <- df_fish %>% 
  group_by(species) %>% 
  summarise(n = n_distinct(survey.seq.no), .groups = "drop") %>%
  mutate(
    sport_flag = case_when(
      species %in% sport.spp ~ "Sportfish", 
      TRUE ~ "Nongame")) 

spp_prev %>% 
  ggplot(aes(x = reorder(species, n), y = n, fill = sport_flag)) + 
  coord_flip() + 
  geom_bar(stat = "identity") + 
  geom_text(
    aes(label=n), 
    position=position_dodge(width=1), hjust=-.2,
    color = "grey20", size = 2) +
  scale_y_continuous(limits = c(0,15500), expand = c(0,0), breaks = seq(0,15000,5000)) + 
  scale_fill_brewer(palette = "Paired") + 
  labs(
    x = "Species common name", 
    y = "Prevelance (count of survey where present)",
    fill = "",
    title = glue::glue("Species (n={n_distinct(df_fish$species)}) prevelance in survey data from 1994 to 2021")) + 
  theme_classic() +
  theme(
    legend.position = c(.7, .9)
  )

# ggsave(filename = here::here("plots","explore","spp_prevelance.pdf"), 
#        dpi = 600, width = 8, height = 16, device = cairo_pdf)
```

```{r pa-matricies}
# P/a for All species ------
pa_all <- 
  df_fish %>% 
  group_by(species, reach_id) %>% 
  summarise(present = sum(fish.count), 
            .groups = 'drop') %>% 
  mutate(present = if_else(present >= 1, 1, 0)) %>% 
  pivot_wider(names_from = "species", values_from = "present", values_fill = 0) %>%
  column_to_rownames("reach_id")

prev_1000 <- spp_prev %>% arrange(desc(n)) %>% slice(1:20) %>% pull(species)

# Filter fish data for spotfish and make P/A matrix
pa_sport <- 
  df_fish %>% 
  filter(species %in% sport.spp) %>% 
  mutate(
    species = case_when(
      species %in%catfish.spp~ "catfish",
      species %in% trout.spp ~ "trout",
      species %in% pan.spp ~ "panfish",
      TRUE ~ species
  )) %>%
  group_by(species, reach_id) %>% 
  summarise(present = sum(fish.count), 
            .groups = 'drop') %>% 
  mutate(present = if_else(present >= 1, 1, 0)) %>% 
  pivot_wider(names_from = "species", values_from = "present", values_fill = 0) %>%
  column_to_rownames("reach_id")
```

<br>

### Principal Components Analysis

We do a PCA on the P/A matrix because it will help us interpret the visualizations from the k-means analysis. 

```{r compute-pca-fish, cache=TRUE}
pca_fish_all <- prcomp(pa_all, center = TRUE, scale. = TRUE)
pca_fish_sport <- prcomp(pa_sport, center = TRUE, scale. = TRUE)
```

```{r pca-fish-plots, cache=TRUE}
# Full biplot
p.pca.fish.all <- 
  fviz_pca_biplot(
    pca_fish_all, 
    geom.ind = c("point"), 
    title = "PCA biplot: P/A all species"
    )

# Variable loadings only
# p.pca.fish.all.vl <- fviz_pca_var(pca_fish_all, col.var = "steelblue")

fviz_pca_var(pca_fish_all, col.var = "steelblue")
 
fviz_pca_var(pca_fish_all, col.var = "contrib", 
   gradient.cols = c("white", "blue", "red"))

fviz_pca_var(pca_fish_all, col.var = "contrib", 
   gradient.cols = c("white", "blue", "red"),
   select.var = list(contrib = 25))

p.pca.fish.all.vl <- 
  fviz_pca_var(pca_fish_all, col.var = "contrib", 
   gradient.cols = c("white", "blue", "red"),
   select.var = list(name = sport.spp))

p.pca.fish.all.vl <- 
  fviz_pca_var(pca_fish_all, col.var = "contrib", 
   gradient.cols = c("white", "blue", "red"),
   select.var = list(name = prev_1000))


# Full biplot
p.pca.fish.sport <- 
  fviz_pca_biplot(
    pca_fish_sport, 
    geom.ind = c("point"), 
    title = "PCA biplot: P/A sportfish species"
    )

# Variable loadings only
p.pca.fish.sport.vl <- fviz_pca_var(pca_fish_sport, col.var = "steelblue")


```


```{r plot-pca-biplot-fish, cache=TRUE, fig.width=12, fig.height=6}
p.pca.fish.all | p.pca.fish.sport
```

```{r plot-pca-vl-fish, cache=TRUE, fig.width=14, fig.height=8}
p.pca.fish.all.vl | p.pca.fish.sport.vl
```

<br>

### K-mean Cluster Analysis

```{r compute-kmeans-fish, cache=TRUE}
# All species
pa_all_clusters <- seq_along(1:10) %>% map(f_kmeans, dat = pa_all)
pa_all_plots <- map(pa_all_clusters, f_fviz_cluster_plots, dat = pa_all)
p.wss.fish.all <- f_wss_plot(10, pa_all_clusters) 

# Sportfish
pa_sport_clusters <- seq_along(1:10) %>% map(f_kmeans, dat = pa_sport)
pa_sport_plots <- map(pa_sport_clusters, f_fviz_cluster_plots, dat = pa_sport)
p.wss.fish.sport <- f_wss_plot(10, pa_sport_clusters)
```

#### Elbow plots for optimal clusters

```{r elbow-fish-panel, fig.height=2, fig.width=8}
p.wss.fish.all | p.wss.fish.sport
```

* The **elbow analyses** suggests optimal grouping for streams based on fish communities is between 2-5 clusters. But there is substantial overlap. For the whole community, there is a good breakpoint at 3 clusters with a steady decline thereafter. For the sportfish communities, there is a good break at 2 groups, then at 4 groups; with little improvement after 6 groups. 

<br>

#### Vizualize clusters: all species (left panels) vs. sportfish (right panels)

```{r plot-clusters-fish-panel, cache=TRUE, fig.width=8, fig.height=10}
(pa_all_plots[[2]] /
   pa_all_plots[[3]] /
   pa_all_plots[[4]] /
   pa_all_plots[[5]]) | 
  (pa_sport_plots[[2]] /
     pa_sport_plots[[3]] /
     pa_sport_plots[[4]] /
     pa_sport_plots[[5]])
```

#### Viz all species clusters

##### 2 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.all.vl | pa_all_plots[[2]]
```


##### 3 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.all.vl | pa_all_plots[[3]]
```


##### 4 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.all.vl | pa_all_plots[[4]]
```


##### 5 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.all.vl | pa_all_plots[[5]]
```


#### Vizualize sportfish clusters

##### 2 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.sport.vl |pa_sport_plots[[2]]
```

* **2 clusters** breaks out streams along a trout-panfish gradient. When we look at the whole community PCA, this may relate to a lotic vs. lentic gradient, which would also suggest some temperature differences.

##### 3 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.sport.vl |pa_sport_plots[[3]]
```

* **3 clusters** starts to break out the streams in the non-trout-stream groups. Perhaps into 1) transitional streams, 2) larger rivers, 3) trout streams. 

##### 4 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.sport.vl |pa_sport_plots[[4]]
```

* **4 clusters** looks to partition out a few groups (with overlap)
  * 1: the bigger river fish (catfish, walleye, sauger and sturgeon), 
  * 2-3: transitional: pike, largemouth, and panfish streams
  * 4: trout streams

##### 5 clusters

```{r cache=TRUE, fig.width=10, fig.height=5}
p.pca.fish.sport.vl |pa_sport_plots[[5]]
```

* **5 clusters** looks to partition out a few groups (with overlap)
  * 1: the bigger river fish (catfish, walleye, sauger and sturgeon), 
  * 2: pike, largemouth, and panfish streams
  * 3-4: transitional?
  * 5: trout streams


<br>

## Environmental & Climate Data

### Data summary

```{r print-env-data}
# skimr::skim(whd_data)
nrow(whd_data)
names(whd_data)
```

### Clean data

We will remove NAs, scale data, and make 2 datasets: 

* one with stream that have fish surveys on them
* one with stream that have fish surveys on them & selected variables

```{r prep-env-data}
# all data minus the distance measures that have too many NAs
whd_data_all <- whd_data %>% 
  select(!ends_with("dist")) %>%
  na.omit() %>% 
  mutate(across(-reach_id, scale)) %>% 
  column_to_rownames("reach_id") 

# stream with fish data
whd_data_fish <- whd_data_all %>% 
  rownames_to_column("reach_id") %>% 
  filter(reach_id %in% df_fish$reach_id)  %>% 
  column_to_rownames("reach_id") 

# stream with fish data and selected temp/flow metrics
whd_data_fish_slct <- whd_data_fish %>% 
  select(trw_area, gradient, sinuosity,
    w_prcp_ann, trw_prcp_ann,
    temp_max_cl_cc, temp_july_cl_cc, temp_summer_cl_cc,
    apr_e10_c, aug_e50_c, wy_e50_c, wy_e90_c | contains("temp"))
```

<br>

### Principal Components Analysis

Again, we do a PCA on the data because it will help us interpret the visualizations from the k-means analysis. 

```{r compute-pca-env, cache=TRUE}
# run analysis
pca_whd_all <- prcomp(whd_data_all, center = TRUE, scale. = TRUE)
pca_whd_fish <- prcomp(whd_data_fish, center = TRUE, scale. = TRUE)
pca_whd_fish_slct <- prcomp(whd_data_fish_slct, center = TRUE, scale. = TRUE)
```

```{r make-pca-plots-env, cache=TRUE}
# Full biplot
p.pca.whd.fish.bi <- 
  fviz_pca_biplot(
  pca_whd_fish, 
  geom.ind = c("point"), 
  title = "Biplot: Streams with fish surveys"
  )

# Variable loadings only
p.pca.whd.fish.vl <- 
  fviz_pca_var(
  pca_whd_fish, 
  col.var = "black", 
  title = "Variable Loadings: Streams with fish surveys",
  repel = TRUE
  )

# Full biplot
p.pca.whd.fish.slct.bi <- 
  fviz_pca_biplot(
  pca_whd_fish_slct, 
  geom.ind = c("point"), 
  title = "Biplot: Streams with fish surveys (select variables)"
  )

# Variable loadings only
p.pca.whd.fish.slct.vl <- 
  fviz_pca_var(
  pca_whd_fish_slct, 
  col.var = "black", 
  title = "Variable Loadings: Streams with fish surveys (select variables)",
  repel = TRUE
  )
```

```{r pca-whd-plot-panel, fig.width=12, fig.height=10}
(p.pca.whd.fish.bi / p.pca.whd.fish.slct.bi) |
  (p.pca.whd.fish.vl / p.pca.whd.fish.slct.vl) 
```

Describe variation in stream attributes. Any gradients?


<br>

### K-means Cluster Analysis

```{r compute-kmeans-env, cache=TRUE}
# All variables
whd_data_fish_clusters <- 
  seq_along(1:10) %>% 
  map(f_kmeans, dat = whd_data_fish)

whd_data_fish_plots <- 
  map(whd_data_fish_clusters, f_fviz_cluster_plots, dat = whd_data_fish)

p.wss.whd.fish <- 
  f_wss_plot(10, whd_data_fish_clusters)


# Reduced Variables
whd_data_fishS_clusters <- 
  seq_along(1:10) %>% 
  map(f_kmeans, dat = whd_data_fish_slct)

whd_data_fishS_plots <- 
  map(whd_data_fishS_clusters, f_fviz_cluster_plots, dat = whd_data_fish_slct)

p.wss.whd.fishS <- 
  f_wss_plot(10, whd_data_fishS_clusters)
```

```{r make-cluster-maps, cache=TRUE}
# All vars
stream_clusters <- 
  whd_data_fish_clusters %>% 
  map(f_get_clusters) %>% 
  map_df(enframe, name = "reach_id", .id = 'k', value = "group")


p.map3 <- f_plot_maps(stream_clusters, 3)
p.map4 <- f_plot_maps(stream_clusters, 4)
p.map6 <- f_plot_maps(stream_clusters, 6)
p.map8 <- f_plot_maps(stream_clusters, 8)


# Reduced vars
stream_clustersS <- 
  whd_data_fishS_clusters %>% 
  map(f_get_clusters) %>% 
  map_df(enframe, name = "reach_id", .id = 'k', value = "group")


p.mapS3 <- f_plot_maps(stream_clustersS, 3)
p.mapS4 <- f_plot_maps(stream_clustersS, 4)
p.mapS6 <- f_plot_maps(stream_clustersS, 6)
p.mapS8 <- f_plot_maps(stream_clustersS, 8)
p.mapS10 <- f_plot_maps(stream_clustersS, 10)
```

<br>

#### Elbow plots for optimal clusters:

```{r elbow-env-panel, fig.height=2, fig.width=8}
p.wss.whd.fish | p.wss.whd.fishS 
```

* The **elbow analyses** looks like a solid breakpoint at 3-4 clusters (sportfish). But still finding reasonable groups at higher resolutions. 


#### Cluster vizualizations (all variables)

```{r plot-clusters-all, cache=TRUE, fig.width=8, fig.height=12}
(whd_data_fish_plots[[3]] / 
   whd_data_fish_plots[[4]] / 
   whd_data_fish_plots[[6]] / 
   whd_data_fish_plots[[8]]) |
  (p.map3 / p.map4 / p.map6 / p.map8)
```


#### Cluster vizualizations (reduced variables)

```{r plot-clusters-slct, cache=TRUE, fig.width=8, fig.height=12}
(whd_data_fishS_plots[[3]] / 
   whd_data_fishS_plots[[4]] / 
   whd_data_fishS_plots[[6]] / 
   whd_data_fishS_plots[[8]]) |
  (p.mapS3 /p.mapS4 / p.mapS6 / p.mapS8)

((whd_data_fishS_plots[[8]]) | (p.mapS8)) 

ggsave(filename = here::here("plots","explore","env_8.pdf"),
       dpi = 600, width = 10, height = 5, device = cairo_pdf)

```


#### Compare: all varibles (left panels) vs. select variables (right panels)

```{r compare-cluster-maps, cache=TRUE, fig.width=8, fig.height=14}
(p.map3 / p.map4 / p.map6 / p.map8) |
  (p.mapS3 /p.mapS4 / p.mapS6 / p.mapS8)
```

<br> 

#### Summary / Interpretations

* things get interesting with using the selected variables vs all variables. Looking at the 8 cluster map comparison, the selected variables look to really get at the heterogeneity of streams within regions. 

```{r, cache=TRUE, fig.width=14, fig.height=5}
p.pca.whd.fish.slct.vl | (whd_data_fishS_plots[[8]]) | (p.mapS8)
```


<br> 

*** 

## Session Info

```{r convert-pdfs, include=FALSE, eval=FALSE}
## convert PDF to PNG
path <- here::here("plots", "explore")
pdfs <- list.files(path, pattern = "*.pdf", recursive = TRUE)
for(pdf in pdfs) {
  pdf_convert(pdf = glue::glue("{path}/{pdf}"), 
              filenames = glue::glue("{path}/{str_remove(pdf, '.pdf')}.png"),
              format = "png", dpi = 500)
}
```

```{r session}
Sys.time()
# git2r::repository()
sessionInfo()
```